Supplementary materials for: Modelling palaeoecological time series using generalized additive models

### 2018-08-16 supplementary-materials.Rmd
mass rebuild after switching to gratia package

---
title: "Supplementary materials for: Modelling palaeoecological time series using generalized additive models"
author: Gavin L. Simpson
affiliation: "Institute of Environmental Change and Society, \\ University of Regina, Regina, Saskatchewan, Canada, S4S 0A2"
mail: "gavin.simpson@uregina.ca"
date: August 15, 2018
keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
bibliography: references.bib
csl: frontiers-in-ecology-and-evolution.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
fontsize: 12pt
mainfont: TeX Gyre Pagella
output:
  pdf_document:
    highlight: tango
    number_sections: true
    fig_width: 8
    fig_height: 5
    fig_crop: false
    keep_tex: true
    latex_engine: xelatex
    md_extensions: +header_attributes+superscript+subscript
    includes:
      in_header: header.tex
  md_document:
    variant: markdown_github
---

@@

```{r load-packages, cache = FALSE}
library("mgcv")
library("scam")
library("ggplot2")
library("cowplot")
library("tidyr")
```
In addition, the *schoenberg* package is required; this in-development package is not on CRAN but can be installed directly from GitHub using functions from the *devtools* package. To install the package, install *devtools* and then use `devtools::install_github()` to install *schoenberg*, as shown below:

```{r install-schoenberg, eval = FALSE}
## schoenberg is not on CRAN, install from github:
install.packages("devtools")
devtools::install_github("gavinsimpson/schoenberg")
```
Assuming that the installation of *schoenberg* completes without error, the package can be loaded as usual
```{r load-schoenberg, cache = FALSE}
library("schoenberg") # need to change the name of the package
```

@@

## Generate a plot of the data
braya_plt <- ggplot(braya, aes(x = Year, y = UK37)) +
    geom_line(colour = "grey") +
    geom_point() +
    labs(y = braya_ylabel, x = "Year CE")

@@

The CAR(1) process plotted in Figure 10 of the manuscript was prepared using

@@

The same model is fitted to the Braya-Sø data set. Note however that in order to even fit the model with both a smooth and the CAR(1) process, I have had to change the default optimiser used to estimate the model parameters, and reduce the basis dimension to a small number. We also fit the model using GCV, which is the defaut, hence no `method` argument

@@

For Braya-Sø, we repeat the process, but we do so for both models (GAMM + CAR(1) and GCV), and use a critical value from the $t$ distribution to form the confidence interval

@@

Figure 5 in the manuscript was produced using:

```{r manuscript-fig-2, fig.show = "hide"}
plot_grid(small_fitted, braya_fitted, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

@@

Clearly we are not able to identify both the wiggly trend and the CAR(1) process in a single model. We might proceed by fitting a simple GAM via `gam()` with REML smoothness selection and moderate number of basis functions --- here we set `k = 15` for illustration.

@@

In the output, the column labelled `k'` is the size of the basis used (the number of basis functions) --- note that it is 14 and not the requested 15 functions as the constant basis function has been removed as it is confounded with the model constant term, the intercept. The column labelled `edf` indicates the *effective degrees of freedom* for the estimated smooth. The `k-index` column contains the value of a test statistic for a test of sufficient number of basis functions --- ideally, we'd like the value of `k-index` to be close to or greater than 1. The quoted *p* value measures the support for the null hypothesis that enough basis functions were available. In practial terms, the low `k-index` and very low *p* value suggest that the requested number of basis functions used to fit the model was too low.

The solution is to increase `k`, say by doubling the original value, and refit the model and perform the basis dimension check using `gam.check()`, repeating this process until `k-index` is close to or greater than 1 and the *p* value is larger than a desired threshold. In my experience, it may not be possible with some data sets to increase `k` large enough to result in a `k-index` > 1 and a *p* value > 0.05, given the available. It is important to remember that the test provided by `gam.check()` is only an heuristic one and shouldn't be seen as infallible. If you find yourself increasing `k` to large values relative to the number of samples in your data set without achieving a `k-index` > 1 or *p* > 0.05, then note the value in the `edf` column as you increase `k`. If the `edf` of the model changes little as you continue to increase `k`, this may be evidence that the test is failing in this particular instance and that the estimated smooth is not dependent on the value of `k` chosen.

@@

To proceed with the Braya-Sø example, we need to increase the basis dimension (`k = 45`), fit using `method = "REML"`. As discussed in the manuscript, we should also account for the non-constant variance, or *heteroscedasticity*, that may arise due to variation in the amount of time (or "lake years") that each sample represents. All else equal, we expect that samples that average more time have smaller variance than those that average a smaller amount of time. To include information about the expected heteroscedasticity in the GAM we can use observational weights[^2]. Here I use the `sampleInterval` variable that I created earlier as the measure of lake years per sample, and to avoid changing the model likelihood, the weights use are the values of `sampleInterval` divided by the mean of `sampleInterval`:

```{r final-braya-reml-fit}
braya_reml <- gam(UK37 ~ s(Year, k = 45), data = braya,
                  method = "REML",
                  weights = sampleInterval / mean(sampleInterval))
summary(braya_reml)
```

@@

If we now check if 44 (`k = 45`) basis functions is sufficient
```{r final-braya-gam-check, fig.show = "hide"}
gam.check(braya_reml)
```
we note that the value of `k-index` is now greater than 1 and the *p* value suggests there is very little evidence against the null hypothesis. Also note that the `edf` of the estimated smooth is considerably below the maximum (`k'`). This is the effect of the smoothness penalty removing the excessive wiggliness possible with 44 basis functions. If the `edf` were close to `k'`, you might consider increasing `k` by a modest amount (in this instance by 5 or 10 additional basis functions) to assure yourself that you have sufficient basis functions. This has the additional advantage of allowing a richer set of smooths of a given complexity (`edf`) to be represented using the basis functions, which may improve the model fit to the data without changing the `edf` too much.

@@

Samples from the posterior distribution of a GAM can be drawn using the `simulate()` methods from the *schoenberg* package.

@@

```{r braya-posterior-simulation}
## posterior simulation
## need to reset-up newBraya
newBraya <- with(braya,
                 data.frame(Year = seq(min(Year), max(Year),
                                       length.out = N)))
braya_pred <- cbind(newBraya,
                    data.frame(predict(braya_reml, newBraya,
                                       se.fit = TRUE)))

## simulate
set.seed(1)
sims2 <- simulate(braya_reml, nsim = nsim, newdata = newBraya,
                  unconditional = TRUE)
colnames(sims2) <- paste0("sim", seq_len(nsim))
sims2 <- setNames(stack(as.data.frame(sims2)),
                      c("simulated", "run"))
sims2 <- transform(sims2, Year = rep(newBraya$Year, nsim),
                       simulated = simulated)

@@

Figure 7 in the manuscript was prepared using

```{r figure-7, fig.show = "hide"}
plot_grid(smallSim.plt, brayaSim.plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

@@

Across-the-function and simultaneous confidence intervals are computed using the `confint()` method. The type of interval required is given via the `type` argument with options `"confidence"` and `"simultaneous"`. Fro example, for Small Water we would use

@@

Figure 8 in the manuscript was prepared using
```{r figure-8, fig.show = "hide"}
plot_grid(smallInt.plt, brayaInt.plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

@@

The first derivative of the estimated trend is calculated using finite differences using the `fderiv()` function. There is also a `confint()` method for objects produced by `fderiv()`. The first derivatives and a 95% simultaneous confidence interval for the Small Water trend we computed and plotted using

@@

Figure 9 in the manuscript was prepared using

```{r figure 9}
plot_grid(small_deriv_plt, braya_deriv_plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

@@

For the Gaussian process smooth to fit within the GAM framework described in the manuscript, we need to supply the value of $\phi$ for the effective range of the correlation function. To estimate $\phi$, we need to repeatedly fit the required GAM using a range of plausible values for $\phi$, which we do using a loop. In the chunk below I fit 200 models with $\phi$ in the range 15--500. For value of $\phi$ I fit the required GAM and extract the REML score (from component `gcv.ubre`) and store it in the numeric vectors `Mat` or `SEx`, for Matérn and Squared Exponential correlation functions, respectively. The final line of the chunk prepares the REML scores for the two correlation functions in long or tidy format suitable for plotting with *ggplot2*.

@@

whilst Figure 12 in the manuscript was prepared using

```{r figure-12, fig.show = "hide"}
plot_grid(proflik.plt, gp.plt2, ncol = 1, labels = c("a","b"),
          align = "hv", axis = "lr")
```

@@

For the model comparison, I refitted all the models for consistency; the code to fit each of the

@@

We write a small function to predict from each model over the range of `Year` and return the data in tidy format for plotting. The plot produce reproduces figure 13 in the manuscript.

@@

allFits <- ggplot(pltData, aes(x = Year, y = fit)) +
    geom_point(aes(x = Year, y = UK37), data = braya) +
    geom_line(aes(colour = Model)) + labs(y = braya_ylabel, x = "Year") +
    theme(legend.position = "right") +
    scale_colour_manual(name = "",
                        values = c("#e66101", "#fdb863", "#5e3c99"))
allFits
```

@@

To make this section self-contained, I refitted the Small Water GAM plus CAR(1) model

@@

and then we load the ^210^Pb dating results for the dated core sections.

@@

The next chunk does the final step in the process. For each of the models we just fitted to include age model uncertainty, we simulate 50 draws from the model posterior distribution. We start with a wrapper function around the `simulate()` code we want to run on each model, then do the actual posterior draws for each model using `lapply()`. The final step just arranges data for plotting.

@@

```{r small-scam-fit-plots, dependson = -1}
plt1 <- ggplot(swAge, aes(y = Date, x = Depth)) +
    geom_line(data = newSims,
              mapping = aes(y = Age, x = Depth, group = Simulation),
              alpha = 1, colour = "grey80") +
    geom_line(data = newAge, mapping = aes(y = fitted, x = Depth)) +
    geom_point(size = 1.5, colour = "red") +
    geom_errorbar(aes(ymin = Date - Error, ymax = Date + Error, width = 0),
                  colour = "red") +
    labs(y = "Year CE", x = "Depth (cm)")

plt2 <- ggplot(simTrends, aes(x = Year, y = Trend, group = Group)) +
    geom_line(alpha = 0.1, colour = "grey80") +
    geom_line(data = newYear,
              mapping = aes(x = Year, y = fit), inherit.aes = FALSE) +
    geom_point(data = small,
               mapping = aes(x = Year, y = d15N),
               inherit.aes = FALSE, size = 0.7) +
    labs(x = "Year", y = d15n_label)

plt3 <- ggplot(simSimulate, aes(x = Year, y = Trend, group = Group)) +
    geom_line(alpha = 0.2, colour = "grey80") +
    geom_point(data = small,
               mapping = aes(x = Year, y = d15N),
               inherit.aes = FALSE,
               size = 0.7) +
    geom_line(data = newYear,
              mapping = aes(x = Year, y = fit),
              inherit.aes = FALSE) +
    labs(x = "Year", y = d15n_label)

@@

This reproduces figure 14 from the manuscript.

>>>>

---
title: "Supplementary materials for: Modelling palaeoecological time series using generalized additive models"
author: Gavin L. Simpson
affiliation: "Institute of Environmental Change and Society, \\ University of Regina, Regina, Saskatchewan, Canada, S4S 0A2"
mail: "gavin.simpson@uregina.ca"
date: August 16, 2018
keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
bibliography: references.bib
csl: frontiers-in-ecology-and-evolution.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
fontsize: 12pt
mainfont: TeX Gyre Pagella
output:
  pdf_document:
    highlight: tango
    number_sections: true
    fig_width: 8
    fig_height: 5
    fig_crop: false
    keep_tex: true
    latex_engine: xelatex
    md_extensions: +header_attributes+superscript+subscript
    includes:
      in_header: header.tex
  md_document:
    variant: markdown_github
---

@@

```{r load-packages, cache = FALSE}
library("mgcv")
library("scam")
library("ggplot2")
library("cowplot")
library("tidyr")
```
In addition, the *gratia* package is required; this in-development package is not on CRAN but can be installed directly from GitHub using functions from the *devtools* package. To install the package, install *devtools* and then use `devtools::install_github()` to install *gratia*, as shown below:

```{r install-gratia, eval = FALSE}
## gratia is not on CRAN, install from github:
install.packages("devtools")
devtools::install_github("gavinsimpson/gratia")
```
Assuming that the installation of *gratia* completes without error, the package can be loaded as usual
```{r load-gratia, cache = FALSE}
library("gratia") # need to change the name of the package
```

@@

## plot Braya-So data
braya_plt <- ggplot(braya, aes(x = Year, y = UK37)) +
    geom_line(colour = "grey") +
    geom_point() +
    labs(y = braya_ylabel, x = "Year CE")

@@

The CAR(1) process plotted in Figure 11 of the manuscript was prepared using

@@

The same model as used for Small Water was also initially attempted with the Braya-Sø data. However, in order to even fit the model with both a smooth and the CAR(1) process, I had to change the default optimizer used to estimate the model parameters, and reduce the basis dimension, `k`, to a small number. I also fit the model using GCV, which is, at the time of writing, the default in `gam()`, hence no `method` argument

@@

To plot the GAMs fitted to the Braya-Sø time series, we repeat the process used with the Small Water GAM, but we do this for both models (GAMM + CAR(1) and GCV), using a critical value from the $t$ distribution to form the confidence interval

@@

Figure 6 in the manuscript was produced using:

```{r manuscript-fig-6, fig.show = "hide"}
plot_grid(small_fitted, braya_fitted, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

@@

For the Braya-Sø data, clearly we are not able to identify both the wiggly trend and the CAR(1) process in a single model. We proceed by fitting a simple GAM via `gam()` with REML smoothness selection and a moderate number of basis functions --- here we set `k = 15` for illustration.

@@

The first few lines of output from `gam.check()` include information on the model fit, whether the algorithm converged, and some diagnostics from the optimization at convergence --- you are unlikely to need this information, but it can be useful in diagnosing problems with model fitting or when reporting problems to Simon Wood, the developer of the *mgcv* package. In the table at the bottom of the output from `gam.check()`, the column labelled `k'` is the size of the basis used to fit the model (the number of basis functions). Note that the stated value is 14 and not the requested 15 functions because the constant basis function has been removed due to it being confounded with the model constant term, the intercept. The column labelled `edf` indicates the *effective degrees of freedom* for the estimated smooth, which is a measure of the complexity or wiggliness of the final smooth. The `k-index` column contains the value of a test statistic for a test of a sufficient number of basis functions --- ideally, we'd like the value of `k-index` to be close to or greater than 1. The quoted *p* value measures the support for the null hypothesis that enough basis functions were available. In practial terms, the low `k-index` and very low *p* value shown above strongly suggest that the requested number of basis functions used to fit the model was too low.

The solution is to increase `k`, say by doubling the original value, and refit the model and perform the basis dimension check using `gam.check()`, repeating this process until `k-index` is close to or greater than 1 and the *p* value is larger than a desired threshold. In my experience fitting GAMs to palaeoenvironmental time series, it may not be possible with some data sets to increase `k` large enough to result in a `k-index` > 1 and a *p* value > 0.05, given the available number of samples. It is important to remember that the test provided by `gam.check()` is only an heuristic one and shouldn't be seen as infallible. If you find yourself increasing `k` to large values relative to the number of samples in your data set without achieving a `k-index` > 1 or *p* > 0.05, then note the value in the `edf` column as you increase `k`. If the `edf` of the model changes little as you continue to increase `k`, this may be evidence that the test is failing in this particular instance and that the estimated smooth is not dependent on the value of `k` chosen.

@@

To proceed with the Braya-Sø example, we need to increase the basis dimension, fit using `method = "REML"`. As discussed in the manuscript, we should also account for the non-constant variance, or *heteroscedasticity*, that may arise due to variation in the amount of time (or "lake years") that each sample represents. All else equal, we expect that samples that average more time have lower variance than those that average a smaller amount of time. To include information about the expected heteroscedasticity in the GAM we can use observational weights[^2]. Here, I use the `sampleInterval` variable that I created earlier as the measure of lake years per sample, and, to avoid changing the model likelihood, the weights use are the values of `sampleInterval` divided by the mean of `sampleInterval`:

```{r final-braya-reml-fit}
braya_reml <- gam(UK37 ~ s(Year, k = 40), data = braya,
                  method = "REML",
                  weights = sampleInterval / mean(sampleInterval))
summary(braya_reml)
```

@@

If we now check if 39 (`k = 40`) basis functions is sufficient
```{r final-braya-gam-check, fig.show = "hide"}
gam.check(braya_reml)
```
we note that the value of `k-index` is now greater than 1 and the *p* value suggests there is very little evidence against the null hypothesis. This result strongly suggests that there were sufficient basis functions used to estimated the trend.

Also note that the `edf` of the estimated smooth is considerably below the maximum (`k'`). This is the effect of the smoothness penalty removing the excessive wiggliness possible with 39 basis functions. If the `edf` were close to `k'`, you might consider increasing `k` by a modest amount (in this instance by 5 or 10 additional basis functions) to assure yourself that you have sufficient basis functions. This has the additional advantage of allowing a richer set of smooths of a given complexity (`edf`) to be represented using this larger set of basis functions, which may improve the model fit to the data without changing the `edf` too much.

@@

Samples from the posterior distribution of a GAM can be drawn using the `simulate()` methods from the *gratia* package.

@@

```{r braya-posterior-simulation}
## data points to simulate at
newBraya <- with(braya,
                 data.frame(Year = seq(min(Year), max(Year),
                                       length.out = N)))
braya_pred <- cbind(newBraya,
                    data.frame(predict(braya_reml, newBraya,
                                       se.fit = TRUE)))

## simulate
set.seed(1)
sims2 <- simulate(braya_reml, nsim = nsim, newdata = newBraya,
                  unconditional = TRUE)

## rearrange the output into a long/tidy format
colnames(sims2) <- paste0("sim", seq_len(nsim))
sims2 <- setNames(stack(as.data.frame(sims2)),
                  c("simulated", "run"))
sims2 <- transform(sims2, Year = rep(newBraya$Year, nsim),
                   simulated = simulated)

@@

Figure 8 in the manuscript was prepared using

```{r figure-8, fig.show = "hide"}
plot_grid(smallSim.plt, brayaSim.plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

@@

Across-the-function and simultaneous confidence intervals are computed using the `confint()` method. The type of interval required is given via the `type` argument with options `"confidence"` and `"simultaneous"`. For example, for Small Water we would use

@@

Figure 9 in the manuscript was prepared using
```{r figure-9, fig.show = "hide"}
plot_grid(smallInt.plt, brayaInt.plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

@@

The first derivative of the estimated trend is calculated using finite differences using the `fderiv()` function. There is also a `confint()` method for objects produced by `fderiv()`. The first derivatives and a 95% simultaneous confidence interval for the Small Water trend were computed and plotted using

@@

Figure 10 in the manuscript was prepared using

```{r figure-10}
plot_grid(small_deriv_plt, braya_deriv_plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

@@

For the Gaussian process smooth to fit within the GAM framework described in the manuscript, we need to supply the value of $\phi$ for the effective range of the correlation function. To estimate $\phi$, we need to repeatedly fit the required GAM using a range of plausible values for $\phi$, which we do using a loop. In the chunk below I fit 200 models with $\phi$ in the range 15--500. For each value of $\phi$ I fit the required GAM and extract the REML score (from component `gcv.ubre`) and store it in the numeric vectors `Mat` or `SEx`, for Matérn and Squared Exponential correlation functions, respectively. The final line of the chunk prepares the REML scores in long or tidy format suitable for plotting with *ggplot2*.

@@

whilst Figure 13 in the manuscript was prepared using

```{r figure-13, fig.show = "hide"}
plot_grid(proflik.plt, gp.plt2, ncol = 1, labels = c("a","b"),
          align = "hv", axis = "lr")
```

@@

For the model comparison I refitted all the models for consistency; the code to fit each of the

@@

I wrote a small function to predict from each model over the range of `Year` and return the data in tidy format for plotting.

@@

allFits <- ggplot(pltData, aes(x = Year, y = fit)) +
    geom_point(aes(x = Year, y = UK37), data = braya) +
    geom_line(aes(colour = Model)) + labs(y = braya_ylabel, x = "Year") +
    theme(legend.position = "right") +
    scale_colour_manual(name = "",
                        values = c("#e66101", "#fdb863", "#5e3c99"))
allFits
```
The plot produced reproduces Figure 14 in the manuscript.

@@

To make this section self-contained, I refitted the Small Water GAM plus CAR(1) model. Because we will be sampling ages for each observation from the posterior of the age-model GAM, the oldest sample will will be somewhat younger than the expected value from the model in some samples. This will cause problems when we come to simulate from the GAMs fitted to the resampled age-models as each GAM will cover a slightly different range of time. To avoid this problem I fix the knots at the extremes of the observed values of `Year` and spread the remaining 12 knots evenly inbetween. The knot locations are stored in the vector `knots`, which is passed to the argument `knots` when fitting the GAM

@@

Setting the knots like this doesn't change the model estimated above; I'm only repeating what happens internally if you don't supply `knots`. However, later we will need to specify this set of knots explicitly when accounting for age model uncertainty.

Next we load the ^210^Pb dating results for the dated core sections.

@@

The next chunk does the final step in the process. For each of the models we just fitted, we simulate 50 draws from the model posterior distribution. We start with a wrapper function around the `simulate()` code we want to run on each model, then do the actual posterior draws for each model using `lapply()`. The final step just arranges data for plotting.

@@

```{r small-scam-fit-plots, dependson = -1}
## plot the estimated age model ad posterior simulations from it
plt1 <- ggplot(swAge, aes(y = Date, x = Depth)) +
    geom_line(data = newSims,
              mapping = aes(y = Age, x = Depth, group = Simulation),
              alpha = 1, colour = "grey80") +
    geom_line(data = newAge, mapping = aes(y = fitted, x = Depth)) +
    geom_point(size = 1.5, colour = "red") +
    geom_errorbar(aes(ymin = Date - Error, ymax = Date + Error, width = 0),
                  colour = "red") +
    labs(y = "Year CE", x = "Depth (cm)")

## plot the simulated trends showing the effect of age-model uncertainty
plt2 <- ggplot(simTrends, aes(x = Year, y = Trend, group = Group)) +
    geom_line(alpha = 0.1, colour = "grey80") +
    geom_line(data = newYear,
              mapping = aes(x = Year, y = fit), inherit.aes = FALSE) +
    geom_point(data = small,
               mapping = aes(x = Year, y = d15N),
               inherit.aes = FALSE, size = 0.7) +
    labs(x = "Year", y = d15n_label)

## plot simulated trends showing the effect of age-model uncertainty and
## the effect of uncertainty in the estimated trend itself
plt3 <- ggplot(simSimulate, aes(x = Year, y = Trend, group = Group)) +
    geom_line(alpha = 0.2, colour = "grey80") +
    geom_point(data = small,
               mapping = aes(x = Year, y = d15N),
               inherit.aes = FALSE,
               size = 0.7) +
    geom_line(data = newYear,
              mapping = aes(x = Year, y = fit),
              inherit.aes = FALSE) +
    labs(x = "Year", y = d15n_label)

@@

This reproduces Figure 15 from the manuscript.
